---
title: "Summary Statistics 1"
author: "BIOE 498/598 PJ"
date: "Spring 2021"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Loading our Birth-day-of-month Data

```{r cars}
days <- as.numeric(readLines("birthdays.txt"))
hist(days, n=31, col="blue", xlab="birth day of month (BDotM)",
     main=sprintf("Population: Mean=%2.2f, Std Dev=%1.2f", mean(days), sd(days)))
```

## Aside: Is this what we expected?

We sampled 30 students. How many do we expect to observe on any day? (Assuming no end-of-month effects.)
```{r echo=TRUE}
lambda = 30 / 31
```

Assuming the underlying process is Poisson, what is the expected frequency?
```{r echo=TRUE}
round(dpois(0:3, lambda), digits=2)
```

Let's count of the observed frequencies.
```{r echo=TRUE}
table(tabulate(days)) / 30
```


## Estimating the mean BDotM

```{r echo=TRUE}
s = sample(days, 5, replace=TRUE)
s
mean(s)
```

## Repeated sampling to estimate the mean BDotM

```{r}
par(mfrow=c(2,2))
for (n in c(10, 100, 1000, 10000)) {
  ss <- replicate(n, mean(sample(days, 5, replace=TRUE)))
  hist(ss, n=30, xlim=c(1,31), xlab="Mean BDotM based on 5 observations", main=paste(n, "samples"))
}
```

## Sampling distribution depends on the \# of observations

```{r}
par(mfrow=c(2,2))
n <- 10000
for (k in c(5, 10, 15, 20)) {
  ss <- replicate(n, mean(sample(days, k, replace=TRUE)))
  hist(ss, n=30, xlim=c(1,31), 
       xlab=paste("Mean BDotM based on", k, "observations"), 
       main=sprintf("%i observations/sample, Std Dev=%1.2f", k, sd(ss)))
}
```