---
title: "The Bootstrap and Distributions"
author: "BIOE 498/598 PJ"
date: "Spring 2021"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
set.seed(56757)
```

## Does BDotM differ for the BIOE 498 and BIOE 598 students?

```{r}
days498 <- c(3,19,15,10,22,23,6,2,2,17,25,15,17,26,13,8,17,10)
days598 <- c(27,16,22,26,7,11,26,16,5,15,7,5)

hist(days498, n=31, col=rgb(0,0,1,0.5), xlab="birth day of month (BDotM)", main="BDotM: 498 (blue) vs. 598 (red)")
hist(days598, n=31, col=rgb(1,0,0,0.5), add=TRUE)
```

## Goal: Test if the difference between 498 and 598 BDotM's is signficant

```{r echo=TRUE}
mean(days498)
mean(days598)
test_diff <- mean(days498) - mean(days598)
test_diff
```

## Creating a boostrap null distribution

```{r echo=TRUE}
# put all the BDotMs in one pool
days <- c(days498, days598)
n498 <- length(days498)
n598 <- length(days598)

diffs <- replicate(10000, {
  sample498 <- sample(n498, days, replace=TRUE)
  sample598 <- sample(n598, days, replace=TRUE)
  # return the difference between the two groups
  mean(sample498) - mean(sample598)
})
```

## Creating a boostrap null distribution

```{r}
plot(density(diffs), main="Bootstrap Null Distribution", xlab="mean(498 BDotM) - mean(598 BDotM)")
abline(v=test_diff, col="red", lty="dashed")
```

## Estimating the $p$-value

The $p$-value is the probability that a difference at least as large can be seen randomly.

We can estimate this probability as the fraction of bootstrap samples that are as large as the test difference.

```{r echo=TRUE}
mean(abs(diffs) >= abs(test_diff))
```

With a $p$-value this large, we cannot reject the null hypothesis that the BDotM is the same for both groups.

## Shortcut method: the $t$-test

```{r echo=TRUE}
t.test(days498, days598, alternative="two.sided")
```

## Shortcut method: the $t$-test (equal variance)

```{r echo=TRUE}
t.test(days498, days598, alternative="two.sided", 
       var.equal=TRUE)
```

## Is the sampling distribution normal?

```{r echo=TRUE}
# Shapiro-Wilk in R is limited to 5000 points
shapiro.test(diffs[1:5000])
```

We reject the null hypothesis of normality.

## Is the sampling distribution normal?

```{r}
car::qqPlot(diffs, id=FALSE, cex=0.3)
```
