---
title: "Introduction to R"
author: "BIOE 498/598 PJ"
date: "1/21/2022"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Everything is a vector

Vectors are built with the `c()` function (for "combine")
```{r}
a <- c(10, 20, 30, 40)
a
```

Vectors are 1-indexed (like Matlab, Fortran, or Julia)
```{r}
a[1]
a[2:4]
```

## Some R artifacts

For historical reasons, R assigns objects to variables using `<-` (said "gets")
```{r}
a <- c(10, 20, 30, 40)
```

However, the equals sign also works
```{r}
a = c(11, 22, 33, 44)
```

Also, periods are allowed in variables names
```{r}
a.var.name <- 7
is.na(a.var.name)
```
It's best to avoid this and use underscores (`_`) instead

## Vector Types

Numeric
```{r}
r <- rnorm(4)
r
```

Logical (`TRUE` and `FALSE` can be abbreviated `T` and `F`)
```{r}
r < 0.0
```

Character
```{r}
c <- c("this", "is", "a", "character")
```

## Notes about characters

Everything is a vector of "strings"; there are no individual "letters"
```{r}
s <- "a string"
length(s)
s[1]
s[2]
```

## Notes about characters (continued)

```{r}
nchar("a string")
c <- c("this", "is", "a", "character")
length(c)
nchar(c)
```

## Everything in a vector is the same type

R will convert objects into the same type
```{r}
c(1, 2, "three")
```

For mixed types, use a `list`
```{r}
list(1, 2, "three")
```

## Factors are special vectors

Factors look like regular vectors
```{r}
x1 <- as.factor(c("wt", "ko", "wt"))
x1
```

\pause
But they are stored as integers and a "key" of levels
```{r}
as.integer(x1)
levels(x1)
```

## Functions

Most functions are *vectorized* and operate elementwise
```{r}
a <- runif(4)
a
10*a + 1
sqrt(a) + cos(a)
```

**You can always ask for help: `?sqrt`.**

## Data Frames

Data frames hold tables of data.

```{r}
data <- read.csv("MonkeyThrow.csv")
data
```

## Size of Data Frames

```{r}
nrow(data)
ncol(data)
```

The `length` of a data frame is the number of *columns*.
```{r}
length(data)
```

## What are data frames?

```{r}
head(data, n=5) # just the first 5 rows
```

 - Each column in a data frame is a vector (and must be the same type).
 - Each row contains data of different types

## What types ("classes") of data are in our data frame?

```{r}
lapply(data, class)  # applies `class` to each column
```

## Working with columns

```{r}
data$hat
```

```{r}
log10(data$distance)
```

## Making new columns

```{r}
data$log_dist <- log10(data$distance)
data
```

## Nevermind, let's delete that column

```{r}
data$log_dist <- NULL
data
```

## Visualizing data

```{r}
plot(data$run, data$distance)
```

## We can do better
```{r}
plot(data$run, data$distance, xlab="run number", 
     ylab="distance", cex=2, pch=20)
```

## Libraries
\footnotesize
Many of the functions we use are in libraries. We need to load the library first.
```{r, fig.height=2.5, fig.width=5}
library("doetools") # include `farplot` function
farplot(data, response="distance", factors=c("hand", "hat", "boots"))
```
