---
title: "Linear Models: Main Effects"
author: "BIOE 498/598 PJ"
date: "Spring 2021"
fontsize: 10pt
output: beamer_presentation
---

```{r echo=FALSE}
knitr::knit_hooks$set(mysize = function(before, options, envir) {
  if (before) 
    return(options$size)
})
```

```{r, include=FALSE}
library(tibble)
library(ggplot2)
library(cowplot)
theme_set(
    theme_classic(base_size = 25)
)
```

## Does BDotM differ for the BIOE 498 and BIOE 598 students?

```{r echo=FALSE}
days498 <- c(3,19,15,10,22,23,6,2,2,17,25,15,17,26,13,8,17,10)
days598 <- c(27,16,22,26,7,11,26,16,5,15,7,5)

hist(days498, n=31, col=rgb(0,0,1,0.5), xlab="birth day of month (BDotM)", main="BDotM: 498 (blue) vs. 598 (red)")
hist(days598, n=31, col=rgb(1,0,0,0.5), add=TRUE)
```

## Shortcut method: the $t$-test

```{r echo=TRUE}
t.test(days498, days598, alternative="two.sided")
```

## New technique: Linear Models

Previous approach: Split data into two groups; test for a difference in BDotM

New appraoch: Build a model that predicts BDotM; ask if 498/598 knowledge helps

## Linear Models

$$ y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_n x_n + \epsilon $$

* $y$ is the **response**
* $x_i$ is a **predictor** or **factor**
* $\beta_0$ is the **intercept**
* $\beta_i, i>0$ is a **coefficient**, **effect size**, or **parameter**
* $\epsilon$ is a **residual**

```{r echo=FALSE}
days_data <- data.frame(day=c(days498, days598), 
                        undergrad=c(rep(TRUE, length(days498)), rep(FALSE, length(days598))))
days_data <- days_data[sample.int(nrow(days_data)), ]
row.names(days_data) <- 1:nrow(days_data)
```

## Re-organizing our data into a **data frame**

```{r}
head(days_data)
```

## Working with data frames

Checking the size (\# rows and \# columns)
```{r}
dim(days_data)
```

\pause
Extracting a single entry
```{r}
days_data[12, ]
```

\pause
Or just the day
```{r}
days_data[12,"day"]
```

## Our data frame contains two vectors

```{r}
days_data$day
days_data$undergrad
```

## Building a linear model with `lm`: $y = \beta_0 + \epsilon$

```{r}
lm( days_data$day ~ 1 )
```

\pause
```{r}
mean(days_data$day)
```

## Modeling with a predictor: $y = \beta_0 + \beta_1 x + \epsilon$
```{r}
lm( days_data$day ~ 1 + days_data$undergrad )
```

\pause
```{r}
mean(days498) - mean(days598)
```

## What does this model mean?

We fit the parameters $\beta_0$ and $\beta_1$ in the model
$$\text{day} = \beta_0 + \beta_1 \times \text{undergrad} + \epsilon$$
\pause
Substituting the fitted values:
$$\text{day} = 15.3 - 1.4 \times \text{undergrad} + \epsilon$$

\pause
For undergrads: $\text{day} = 15.3 - 1.4\times1 = 13.9$

\pause
For grad students: $\text{day} = 15.3 - 1.4\times0 = 15.3$

\pause
These are the means for each group.

## Let's clean up our calls to `lm`: Intercepts

An intercept is always assumed.
```{r eval=FALSE}
lm( y ~ 1 + x )
```
is equivalent to
```{r eval=FALSE}
lm( y ~ x )
```

If you don't want an intercept, use a `0`
```{r eval=FALSE}
lm( y ~ 0 + x )
```

## Let's clean up our calls to `lm`: Naming a data frame

We can give `lm` a data frame where it can find our response and predictor variables.
```{r eval=FALSE}
lm( days_data$day ~ days_data$undergrad )
```
is equivalent to
```{r eval=FALSE}
lm( day ~ undergrad, data=days_data )
```

## Storing our model for further analysis

We can assign the output of a model to a variable.
```{r}
model <- lm( day ~ undergrad, data=days_data )
summary(model)
```

## Testing a new blood presure medication
```{r, include=FALSE}
library(tibble)
bp_data <- tribble(
  ~BPchange, ~treated, ~male,
  -0.5252988, TRUE, FALSE,
  4.1742790,  TRUE, FALSE,
  6.0256724, TRUE, TRUE,
  -1.4029857, TRUE, FALSE,
  0.4926655,  TRUE, FALSE,
  12.8885132,  FALSE, TRUE,
  6.7517331, FALSE, TRUE,
  4.9016320, TRUE, FALSE,
  45.95538,  FALSE, TRUE,
  19.05073,  FALSE, FALSE,
  30.32632,  TRUE, TRUE,
  33.18965,  FALSE, TRUE,
  5.73430,  TRUE, FALSE,
  33.07533,  FALSE, TRUE,
  5.58772,  FALSE, FALSE,
  31.44072,  FALSE, TRUE
)
```

```{r}
head(bp_data)
```

## Does our BP treatment work?
```{r}
qplot(data=bp_data, x=treated, y=BPchange, size=I(5))
```

## Hypothesis testing the treatment effect
```{r}
t.test(BPchange ~ treated, data=bp_data)
```


## A linear model with effect of treatment

```{r}
model <- lm(BPchange ~ treated, bp_data)
summary(model)
```

## What else could explain the effect?
```{r}
summary(lm(BPchange ~ treated + male, bp_data))
```

## Does our BP treatment work?
```{r}
qplot(data=bp_data, x=treated, y=BPchange, color=male, size=I(5))
```

## Summary

* Linear models can be used for hypothesis testing.

\pause
* Multivariate linear models consider how **all** factors affect the response. This is a form of conditioning.

\pause
* Next time: What if the factors interact?