---
title: "Calculating Effects with Linear Models"
author: "BIOE 498/598 PJ"
date: "Spring 2022"
output: beamer_presentation
fontsize: 9pt
header-includes: 
  - \usepackage{tikz}
  - \usepackage{booktabs}
  - \usepackage{siunitx}
  - \renewcommand\mathfamilydefault{cmr}
  - \usetikzlibrary{positioning,shapes.misc,calc,backgrounds,scopes} 
  - \newcommand\factor[1]{{\ensuremath{\boldsymbol{\text{#1}}}}}
  - \newcommand\fT{\factor{T}}
  - \newcommand\fC{\factor{C}}
  - \newcommand\fK{\factor{K}}
  - \newcommand\fTC{\factor{TC}}
  - \newcommand\fTK{\factor{TK}}
  - \newcommand\fCK{\factor{CK}}
  - \newcommand\fTCK{\factor{TCK}}
  - \newcommand\hi{\ensuremath{\boldsymbol{+}}}
  - \newcommand\lo{\ensuremath{\boldsymbol{-}}}
  - \newcommand\ME{\ensuremath{\mathit{ME}}}
  - \newcommand\INT{\ensuremath{\mathit{Int}}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## An additive model of effects and interactions

For our reactor example with factors $\fT$, $\fC$, and $\fK$:
\begin{align*}
\text{yield} = \beta_0 &+ \beta_\fT \fT + \beta_\fC \fC + \beta_\fK \fK \\
            &+ \beta_{\fTC}\fTC + \beta_{\fTK}\fTK + \beta_{\fCK}\fCK \\
            &+ \beta_{\fTCK}\fTCK
\end{align*}
where each regression coefficient $\beta_i$ is half of the $i$th effect:
\[ \beta_\fT = \frac{\ME(\fT)}{2}, \ldots, \beta_{\fTCK} = \frac{\INT(\fTCK)}{2} \]
and the intercept $\beta_0$ is the mean of all the responses.

## For the reactor example

\begin{columns}
\begin{column}{0.35\textwidth}
  \begin{center}
    \begin{tabular}{rS[table-format=3.2]S[table-format=3.2]}
      \toprule
      \textbf{Effect} & \textbf{Size} & $\boldsymbol{\beta}$ \\
      \midrule
      intercept & 64.25 & 64.25 \\
      $\fT$ & 23 & 11.5 \\
      $\fC$ & -5 & -2.5 \\
      $\fK$ & 1.5 & .75 \\
      $\fTC$ & 1.5 & .75 \\
      $\fTK$ & 10 & 5 \\
      $\fCK$ & 0  & 0 \\
      $\fTCK$ & .5 & .25 \\
      \bottomrule
    \end{tabular}
  \end{center}
\end{column}

\begin{column}{0.6\textwidth}
  \begin{align*}
    \text{yield} = 64.25 &+ 11.5\, \fT - 2.5 \fC + 0.75\, \fK \\
             &+ 0.75\, \fTC + 5\, \fTK + 0\, \fCK \\
             &+ 0.25\, \fTCK
  \end{align*}
\end{column}
\end{columns}

\pause
\bigskip
What is the yield for the treatment $\fT=\hi$, $\fC=\lo$, $\fK=\hi$?
\pause
\begin{align*}
  \text{yield} &= 64.25 + 11.5(1) -2.5(-1) + 0.75(+1) \\
    &\qquad + 0.75(1)(-1) + 5(1)(1) + 0(-1)(1) \\
    &\qquad + 0.25(1)(-1)(1) \\
    &= 64.25 + 11.5 + 2.5 + 0.75 - 0.75 + 5 + 0 - 0.25 \\
    &= 83
\end{align*}

## Why are the coefficients half of the effect sizes?

Imagine the simplest model with one factor $\fT$.
\[ y = \beta_0 + \beta_\fT \fT \]
\pause
Remember the definiton of the main effect of $\fT$:
\[ \ME(\fT) = \bar{y}(\fT\hi) - \bar{y}(\fT\lo) \]
\pause
From our model
\begin{align*}
  \bar{y}(\fT\hi) &= \beta_0 + \beta_\fT (+1) \\
                  &= \beta_0 + \beta_\fT \\
                  \\
  \bar{y}(\fT\lo) &= \beta_0 + \beta_\fT (-1) \\
                  &= \beta_0 - \beta_\fT
\end{align*}
\pause
So according to the model,
\begin{align*}
  \ME(\fT) &= \bar{y}(\fT\hi) - \bar{y}(\fT\lo) \\
           &= (\beta_0 + \beta_\fT) - (\beta_0 - \beta_\fT) \\
           &= 2\beta_\fT
\end{align*}

## Advantages of estimating effects with linear models

- Easier calculation

- Statistical significance for coefficients (coming soon!)

- Predictions for untested treatments

- Extrapolation beyond the design space (for quantitative factors)

## Step 1: Load the data

```{r}
pilot_data <- read.csv("PilotPlantDesign.csv")
pilot_data
```

## Step 2: Look at the data

We can visualize factorial designs with a factor-and-response plot, or farplot.

First, we need to install the `doetools` package. If you haven't already, install the `devtools` package:
```{r eval=FALSE}
install.packages("devtools")
```
You only need to run this once this semester.

Now you can use `devtools` to install `doetools`:
```{r eval=FALSE}
devtools::install_github("jensenlab/doetools")
```
You should re-install this package before every assignment in case we add anything to the package during the semester.

## Now, let's visualize the data

```{r fig.width=4, fig.height=3}
library(doetools)
farplot(pilot_data, response="yield", factors=c("T", "C", "K"))
```

## Step 3: Fit a linear model

```{r}
model <- lm(yield ~ T * C * K, data=pilot_data)
show_effects(model)
```

## Fitting linear models with `lm`

The standard call to `lm` is
```{r eval=FALSE}
model <- lm(<formula>, data=<dataframe>)
```
where `<formula>` takes the form
```{r eval=FALSE}
response ~ effects
```

\bigskip
\pause
Formulas don't include the coefficients, only the effects. For example the model
\[ \text{yield} = \beta_0 + \beta_\fT \fT + \beta_\fK \fK + \beta_\fTK \fTK \]
would be specified
```{r eval=FALSE}
yield ~ 1 + T + K + T:K
```
\pause
or, since R assumes the model has an intercept, we can omit the `1`.
```{r eval=FALSE}
yield ~ T + K + T:K
```

## More about formulas

The `*` operator is a shortcut for adding main effects and interactions.
```{r eval=FALSE}
yield ~ T + K + T:K
```
is equivalent to
```{r eval=FALSE}
yield ~ T*K
```

\pause
Our complete model
```{r eval=FALSE}
yield ~ T + C + K + T:C + T:K + C:K + T:C:K
```
can be written
```{r eval=FALSE}
yield ~ T*C*K
```

\bigskip
\pause
**The effects in the formula must match column names in the data frame.** R looks up the effects in the data frame to construct the *model matrix*.

## Why do we prefer coded factors?

```{r}
pilot_planning <- read.csv("PilotPlantPlanning.csv")
pilot_planning
model_uncoded <- lm(yield ~ T*C*K, data=pilot_planning)
```

## Comparing coded vs. uncoded models
::: columns
:::: column
```{r}
show_effects(model) # coded
```
::::
:::: column
```{r}
show_effects(model_uncoded)
```
::::
:::

\pause
::: columns
:::: column
```{r echo=FALSE}
pilot_planning$T_fahr <- pilot_planning$T*9/5 + 32
model_fahr <- lm(yield ~ T_fahr*C*K, data=pilot_planning)
```
::::
:::: column
\bigskip
```{r}
show_effects(model_fahr)
```
::::
:::




