---
title: "Active Effects"
author: "BIOE 498/598 PJ"
date: "Spring 2022"
output: beamer_presentation
fontsize: 9pt
header-includes: 
  - \usepackage{tikz}
  - \usepackage{booktabs}
  - \usepackage{siunitx}
  - \renewcommand\mathfamilydefault{cmr}
  - \usetikzlibrary{positioning,shapes.misc,calc,backgrounds,scopes} 
  - \newcommand\factor[1]{{\ensuremath{\boldsymbol{\text{#1}}}}}
  - \newcommand\fT{\factor{T}}
  - \newcommand\fC{\factor{C}}
  - \newcommand\fK{\factor{K}}
  - \newcommand\fTC{\factor{TC}}
  - \newcommand\fTK{\factor{TK}}
  - \newcommand\fCK{\factor{CK}}
  - \newcommand\fTCK{\factor{TCK}}
  - \newcommand\hi{\ensuremath{\boldsymbol{+}}}
  - \newcommand\lo{\ensuremath{\boldsymbol{-}}}
  - \newcommand\ME{\ensuremath{\mathit{ME}}}
  - \newcommand\INT{\ensuremath{\mathit{Int}}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Active effects

- An **active effect** has an effect size that is large enough to be **practically significant**.

- An **inactive effect** (or **inert effect**) does not have a practically significant effect size.

\pause
How do we assess practical significance?

1. Any effect that causes a meaningful change based on our knowledge of the system or process.
    - A 3% increase in yield for a commodity chemical process might be significant financially.
    - A 3% decrease in tumor size after treatment might be insignificant to the patient.
  
2. Lacking insight about the system or process, we can estimate practical significance by *comparing* effect sizes in the same experiment.

## Bioprocess conversion case study

Goal: Improve bioprocess conversion of switchgrass to biofuel

- Four process factors
  - $\factor{S}$: Bacterial strain (strain A or B)
  - $\factor{T}$: Temperature (30 $^\circ$C or 37 $^\circ$)
  - $\factor{M}$: Mineral supplement (no or yes)
  - $\factor{R}$: Stirring rate (fast or slow)
  
- Response: Percent conversion of carbon
  
- $2^4$ factorial study = 16 runs

- Unreplicated design

- Randomized run order

## Step 1: Load the data

```{r}
pds <- read.csv("ProcessDevelopmentStudy.csv")
head(pds)
```

## Step 2: Look at the data

First, let's check if the response is correlated with run order.
```{r fig.width=4, fig.height=3}
plot(pds$run, pds$conversion)
```

## Step 2: Look at the data

```{r fig.width=4, fig.height=3}
library(doetools)
farplot(pds, response="conversion", factors=c("S","T","R","M"))
```

## Step 3: Fit a linear model

```{r}
model <- lm(conversion ~ S*T*R*M, data=pds)
show_effects(model, scaling=2)
```

## Ordering the effects by magnitude

```{r}
show_effects(model, order="abs", scaling=2, intercept=FALSE)
```

## Visualizing effect sizes

```{r fig.width=5, fig.height=3}
effect_dotplot(model, scaling=2, binwidth=0.5)
```

## Permutation tests

```{r}
red <- c(8,3,9,6)
black <- c(7,4,5,2)
test <- mean(red) - mean(black)
test
```

```{r include=FALSE}
set.seed(498598)
```

\bigskip
\pause
```{r}
cards <- c(red, black)
N <- 200
nulls <- numeric(N)
for (i in 1:N) {
  idxs <- sample(1:8, 4)
  nulls[i] <- mean(cards[idxs]) - mean(cards[-idxs])
}
```

## Permutation test results

```{r fig.height=4}
hist(nulls, n=sqrt(N))
abline(v=test, col="red", lwd=2)
```

Calculating a $p$-value from the null distribution.
```{r}
1 - sum(nulls <= test) / length(nulls)
```


## Why can we compare effect sizes in the same model?

- A *permutation test* creates a null distribution by randomly re-assigning data to groups.

- Each contrast in a factorial experiment is a permutation of the responses, so the *inactive* effect sizes create a null distribution.


## A larger ($2^5$) unreplicated study

```{r}
tumor <- read.csv("TumorInhibition.csv")
head(tumor)
```
## Step 2: Look at the data
```{r fig.width=5, fig.height=3}
farplot(tumor, response="inhibition", factors=c("A","B","C","D","E"))
```

## Step 3: Fit a linear model

```{r}
model <- lm(inhibition ~ A*B*C*D*E, data=tumor)
show_effects(model, scaling=2, intercept=FALSE, order="abs", n=18)
```

## Visualizing effect sizes

```{r fig.width=5, fig.height=3}
effect_dotplot(model, scaling=2, binwidth=0.5)
```

## Half-normal plots

- With enough effects, the inactive factors will approximate a normal distribution.

- A *half-normal plot* displays the effect sizes and their associated probabilities.

- Inactive effects fall along a straight line, while active effects deviate.

- The half-normal plot uses the effect *magnitudes* since the sign depends **only** on how the \lo\ and \hi\ levels were assigned.

## Selecting active effects with a half-normal plot

```{r include=FALSE}
library(daewr)
```

```{r results=FALSE, fig.width=6, fig.height=4}
daewr::halfnorm(get_effects(model))
```

\pause
Active effects: factors \factor{B}, \factor{D}, & \factor{E} and interactions \factor{BD} & \factor{DE}.
