---
title: "Alternative Fractional Factorial Designs"
author: "BIOE 498/598 PJ"
date: "Spring 2022"
output: beamer_presentation
fontsize: 9pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(magrittr)
library(doetools)
```

## How low can we go?

\newcommand\lo{\ensuremath{\boldsymbol{-}}}
\newcommand\hi{\ensuremath{\boldsymbol{+}}}
\newcommand\iii{\ensuremath{\mathrm{III}}}
\newcommand\iv{\ensuremath{\mathrm{IV}}}

```{r include=FALSE}
designs <- tibble::tribble(
  ~factors, ~fraction, ~resolution,
  2, 0, NA,
  3, 0, NA,
  3, 1, 3,
  4, 0, NA,
  4, 1, 4,
  5, 0, NA,
  5, 1, 5,
  5, 2, 3,
  6, 0, NA,
  6, 1, 6,
  6, 2, 4,
  6, 3, 3,
  7, 0, NA,
  7, 1, 7,
  7, 2, 4,
  7, 3, 4,
  7, 4, 3,
  8, 0, NA,
  8, 1, 8,
  8, 2, 5,
  8, 3, 4,
  8, 4, 4,
  9, 0, NA,
  9, 1, 9,
  9, 2, 6,
  9, 3, 4,
  9, 4, 4,
  9, 5, 3,
  10, 0, NA,
  10, 1, 10,
  10, 2, 7,
  10, 3, 5,
  10, 4, 4,
  10, 5, 4,
  10, 6, 3
)

designs %<>%
  dplyr::group_by(factors, resolution) %>%
  dplyr::summarise(fraction = max(fraction)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(runs = 2^(factors - fraction)) %>%
  dplyr::filter(resolution <= 5 | is.na(resolution)) %>%
  dplyr::mutate(resolution = ifelse(is.na(resolution), "full", as.character(resolution))) %>%
  dplyr::mutate(resolution = factor(resolution, levels=c("full", "5", "4", "3")))
```

The efficiency of fractional factorial designs offsets the exponential increase in runs for factorial designs.

```{r out.width='75%', fig.align='center', echo=FALSE}
ggplot2::qplot(data=designs, 
               x=factors, y=runs, 
               color=resolution, 
               geom=c("line", "point")) +
  ggplot2::theme_classic(base_size=36)
```

## How low can we go? (zoomed in)

```{r out.width='85%', fig.align='center', echo=FALSE}
ggplot2::qplot(data=designs %>% dplyr::filter(runs < 150), 
               x=factors, y=runs, 
               color=resolution, 
               geom=c("line", "point"),
               ylim=c(0,150)) +
  ggplot2::theme_classic(base_size=36)
```

##
\begin{columns}

\begin{column}{0.8\textwidth}
\begin{center}
\includegraphics{figures/frf2-resolution.png}
\end{center}
\end{column}

\begin{column}{0.2\textwidth}
{\footnotesize Gromping, 2014\\ \emph{J.\ Stat.\ Software}}
\end{column}

\end{columns}

## Foldover Designs

\begin{columns}

\begin{column}{0.5\textwidth}
Imagine a $2^{6-3}_\iii$ design with 
\[ D=AB, \quad E=AC,\quad F=BC \]
\begin{align*}
  I &=ABD=ACE=BCF=DEF \\
  &=BCDE=ACDF=ABEF
\end{align*}

After analysis, we find that both $B$ and $D$ are significant.

\medskip
Since $D=AB$, the significance of $D$ might be due to $B$ and $AB$.

\medskip
We can \emph{augment} the design by doubling the runs \emph{with $D$ flipped}. This clears $D$ and its interactions.
\end{column}

\begin{column}{0.5\textwidth}
\begin{tabular}{ccccccc}
Run & $A$ & $B$ & $C$ & $D$ & $E$ & $F$ \\
\hline
1 & \lo & \lo & \lo & \hi & \hi & \hi \\
2 & \hi & \lo & \lo & \lo & \lo & \hi \\
3 & \lo & \hi & \lo & \lo & \hi & \lo \\
4 & \hi & \hi & \lo & \hi & \lo & \lo \\
5 & \lo & \lo & \hi & \hi & \lo & \lo \\
6 & \hi & \lo & \hi & \lo & \hi & \lo \\
7 & \lo & \hi & \hi & \lo & \lo & \hi \\
8 & \hi & \hi & \hi & \hi & \hi & \hi \\
\hline
\end{tabular}

\pause
\begin{tabular}{ccccccc}
 9 & \lo & \lo & \lo & \lo & \hi & \hi \\
10 & \hi & \lo & \lo & \hi & \lo & \hi \\
11 & \lo & \hi & \lo & \hi & \hi & \lo \\
12 & \hi & \hi & \lo & \lo & \lo & \lo \\
13 & \lo & \lo & \hi & \lo & \lo & \lo \\
14 & \hi & \lo & \hi & \hi & \hi & \lo \\
15 & \lo & \hi & \hi & \hi & \lo & \hi \\
16 & \hi & \hi & \hi & \lo & \hi & \hi \\
\hline
\phantom{Run} &  \\
\end{tabular}
\end{column}

\end{columns}

## Mirror image designs

If we combine a Resolution \iii\ design with its mirror image (all factors flipped), we have a Resolution \iv\ design with all main effects clear.

\medskip
If we add a blocking factor we can perform the experimental batches sequentially.

\medskip
As with foldover designs, mirror image designs are only necessary if more than one main effect is significant.

## Blocked Designs

- Sometimes logistics force us to group runs into "blocks".

- Examples
  - Mice need to be housed in separate cages.
  - The experimenter cannot do all runs in a single day/batch.
  - Two experimenters need to split up the runs.
  - A single container of reagent doesn't cover the experiment.

\pause
- A "blocking factor" is added to these experiment to capture inter-block differences.

- Blocks are added to designs and analyzed like any other factor.

- Most block $\times$ factor interactions are ignored.

## Blocking our Pilot Plant Experiment

```{r fig.width=5, fig.height=3}
pilot <- read.csv("PilotPlantBlocked.csv")
farplot(pilot, "yield", c("T","C","K","block"))
```

## Building our normal model

```{r}
model <- lm(yield ~ T*K*C, data=pilot); show_effects(model)
```

\pause

```{r}
model <- lm(yield ~ T*K*C + block, data=pilot); show_effects(model)
```

