---
title: "Replication"
author: "BIOE 498/598 PJ"
date: "Spring 2022"
output: beamer_presentation
fontsize: 9pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(doetools)
library(daewr)
#source("show_model.R")
```

## Why replication?

1. Reduce noise effects.

2. Estimate confidence intervals for effect sizes.

3. Analyze dispersion effects.

## Does noise matter?

```{r}
data <- read.csv("LeafSpring.csv")
data1 <- data[ 1:16, ]
data2 <- data[17:32, ]
data3 <- data[33:48, ]
```

## Effect sizes differ for each replicate

```{r include=FALSE}
show_eff <- function(d) {show_effects(lm(height ~ B*C*D*E*Q, data=d), ordered="abs", intercept=FALSE, n=10)}
```

::: columns
:::: column
Replicate 1
```{r echo=FALSE}
show_eff(data1)
```
::::
:::: column
Replicate 2
```{r echo=FALSE}
show_eff(data2)
```
::::
:::

## Duplicates vs. Replicates

- We use replicates to estimate **confidence intervals** for effect sizes.

- More replicates = narrower confidence intervals

- Treating duplicates as replicates artificially narrows the confidence intervals for two reasons.
  1. Inflating the degrees of freedom fakes additional statistical power
  2. Duplicates typically have less variation than replicates
  
\pause
- Statistical significance $\ne$ Practical significance

## Sample variance across replicates

If a run is replicated $r$ times with responses $y_1$, $y_2$, $\ldots$, $y_r$ and mean $\bar{y}$,

\[ \text{sample variance} = s^2 = \frac{\sum_i^n (y_i - \bar{y})^2}{r-1} \]

\pause
For a factorial design with $N$ *unreplicated* runs ($N=2^k$ for a full factorial or $N=2^{k-p}$ for a fractional factorial),

\[ \text{standard error of effects} = SE(\beta_i) = \sqrt{\frac{\text{mean}(s^2)}{rN}} \]

## Linear models find the "best fit" effect sizes

```{r}
model <- lm(height ~ B*C*D*E*Q, data=data)
show_model(model, n_coefs=17, show_fit=FALSE)
```
## Half-normal & dot plots --- significance based only on effect size

```{r warning=FALSE, message=FALSE, echo=FALSE}
effect_dotplot(model)
```

```{r echo=FALSE, fig.width=5, fig.height=3, warning=FALSE, message=FALSE}
daewr::halfnorm(na.omit(get_effects(model)))
```


