---
title: "Dispersion"
author: "BIOE 498/598 PJ"
date: "Spring 2022"
output: beamer_presentation
fontsize: 9pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(doetools)
library(daewr)
library(magrittr)
#source("show_model.R")
```

## Why replication?

1. Reduce noise effects.

2. Estimate confidence intervals for effect sizes.

3. Analyze dispersion effects.

```{r include=FALSE}
data <- read.csv("LeafSpring.csv")
data %<>% dplyr::arrange(B, C, D, E, Q) %>% as.data.frame()
```

## Sample variance across replicates

If a run is replicated $r$ times with responses $y_1$, $y_2$, $\ldots$, $y_r$ and mean $\bar{y}$,

\[ \text{sample variance} = s^2 = \frac{\sum_i^r (y_i - \bar{y})^2}{r-1} \]

\pause
For a factorial design with $N$ *unreplicated* runs ($N=2^k$ for a full factorial or $N=2^{k-p}$ for a fractional factorial),

\[ \text{standard error of effects} = SE(\beta_i) = \sqrt{\frac{\text{mean}(s^2)}{rN}} \]

## Visualizing the data
```{r fig.width=5, fig.height=3}
farplot(data, factors=c("B","C","D","E","Q"), response="height")
```

## Linear models find the "best fit" effect sizes

```{r}
model <- lm(height ~ B*C*D*E*Q, data=data)
show_model(model, n_coefs=17, show_fit=FALSE)
```

## Half-normal & dot plots --- significance based only on effect size

```{r warning=FALSE, message=FALSE, echo=FALSE}
effect_dotplot(model)
```

```{r echo=FALSE, fig.width=5, fig.height=3, warning=FALSE, message=FALSE}
daewr::halfnorm(na.omit(get_effects(model)))
```
## Location vs. Dispersion

- Sometimes we want to study the variation in the response, not the response itself.

- **Location** describes the central tendency of a response
  - Mean, median, mode
  - All of our models so far use response = location
  
\pause
- **Dispersion** describes the spread of a response
  - Range, inter-quartile range (IQR), variance, standard deviation
  
\pause
- Location can be studied with unreplicated or replicated designs
- Studying dispersion always requires replicates

## Studying dispersion

- The variance $\sigma^2$ is the natural statistic for studying dispersion with linear models fit by least-squares

- However, the sample variance $s^2$ is not a good response for studying $\sigma^2$
  - $s^2$ is left-censored ($s^2 \ge 0$)
  - $s^2$ follows a $\chi^2$ distribution, not a normal distribution
  
\pause
- Both problems are fixed by modeling $\ln s^2$ instead of $s^2$

\pause
- Moreover, maximizing $-\ln s^2$ minimizes the variance, so we can keep the same maximization-based framework used for location models

## Calculating $\ln s^2$

```{r include=FALSE, eval=FALSE}
add_dispersion <- function(data, factors, response, df=TRUE) {
  data %<>% 
    dplyr::group_by(.dots={{factors}}) %>%
    dplyr::summarize(location__ = mean(.data[[response]]), lns2 = log(var(.data[[response]]))) %>%
    dplyr::ungroup()
  data[[response]] <- data$location__
  data$location__ <- NULL
  if (df) {
    as.data.frame(data)
  } else {
    df  # return a tibble
  }
}
```

```{r}
disp <- add_dispersion(data, factors=c("B","C","D","E","Q"), 
                       response="height")
```
::: {.columns}
:::: {.column width="40%"}
```{r}
head(data, n=16)
```
::::
:::: {.column width="60%"}
```{r}
disp
```
::::
:::

## Visualizing the **dispersion**
```{r fig.width=5, fig.height=3}
farplot(disp, factors=c("B","C","D","E","Q"), response="lns2")
```

## Visualizing the **data**
```{r fig.width=5, fig.height=3}
farplot(data, factors=c("B","C","D","E","Q"), response="height")
```

## Building the model

```{r}
disp_model <- lm(-lns2 ~ B+C+D+E+Q + 
                         B:Q + C:Q + D:Q + E:Q + B:C + B:D + B:E, 
                 data=disp)
```

::: {.columns}
:::: {.column width="40%"}
Confounding in the $2^{5-1}$ design with I=BCDE:

- main effects clear
- BQ
- CQ
- DQ
- EQ
- BC=DE
- BD=CE
- BE=CD
::::
:::: {.column width="60%"}
```{r}
show_effects(disp_model, ordered="abs")
```
::::
:::

## Factors affecting **location** (spring height)
```{r fig.width=5, fig.height=4, warning=FALSE, message=FALSE}
daewr::halfnorm(na.omit(get_effects(model)))
```

## Factors affecting **dispersion** ($\ln s^2$)

```{r fig.width=5, fig.height=4, warning=FALSE, message=FALSE}
daewr::halfnorm(get_effects(disp_model))
```
