---
title: "Homework Review"
author: "BIOE 498/598"
date: "2/12/2020"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## What to show us:

- **Objectives**: what did you want to find out.
- **Data**
  - Where you got it.
  - Ranges and *distributions*.
  - Units
  
## Correlation plots:

Use these for diagnosis; only show anything interesting.
```{r}
data <- read.csv("Melanoma_screen.csv"); plot(data)
```

## Methods

What you did and *why*. No results; keep it brief but precise.

## Results

- State results so they are meaningful for your audience.
  - **Bad**: The effect size for cell_lineSkMel was significant ($\beta=-0.08$, $p<10^-5$).
  - **Good**: The viability for SkMel cells was 8\% lower on average.
- Effect sizes matter; people conflate $p$-values with effect sizes, so it's best to avoid reporting them.
- Only report significant conclusions. This avoids having to say ``significantly" after every result.
- Insignificant results of interest can also be reported but should be labeled as such.

## How to write Results

1. Write down a result.
2. Ask yourself ``what does this mean?"
3. Delete what you originally wrote and write the meaning instead.

## Conclusions

- Results answer ``what does this mean?".
- Conclusions answer ``why does this matter?".
- Example:
  - **Result**: Viability of SkMel cells was 8\% lower on average.
  - **Conclusion**: We need to decide if which cell line is a better model. Other results on SkMel cell lines should be normalized before comparisons are made.

## Now, let's look at our data.

```{r}
plot(sort(data$viability))
```
